A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-01-07 podc_paper/boundedspace.tex
working on GC

Known lock-free search trees have step complexity that includes a term linear in $p$ \cite{EFHR14,Ko20}.  
However, we do not require all the standard search tree functions.
Instead of a standard insertion, we allow a \op{Refresh}'s insertion to fail if another
concurrent \op{Refresh} succeeds in inserting a node, just as the CAS on line \ref{cas}
can fail if a concurrent \op{Refresh} does a successful CAS.
Thus, we can use a particularly simple concurrent RBT implementation.
A sequential RBT can be made persistent using the classic node-copying technique of 
Driscoll et al.~\cite{DSST89}:  all nodes are immutable, and operations on the 
tree are performed by making a new copy of each node $v$ that must be modified, as well
as each node along the path from the root to $v$.
The tree reachable from the new copy of the root is the result of applying the tree operation.
This only adds a constant factor to the running time of any routine designed for a (sequential) RBT.
Once a process has performed an update to the RBT representing the blocks of a node 
$v$ in the \ordering\ tree, 
it uses a CAS to swing $v$'s pointer from the previous RBT root to the new RBT root.
A search in the RBT can simply read the pointer to the RBT root and perform a standard
sequential search on it.
Bashari and Woelfel ~\cite{DBLP:conf/podc/BashariW21} used persistent red-black trees in a similar way to implement a snapshot data structure.

@@

To maintain good amortized time per operation, we periodically do a garbage collection (GC) phase.
If a process performing a \op{Refresh} on the root finds that $\var{root}.\head$ is a multiple
of $p^2$, it does GC.
To determine which blocks can be thrown away, we keep an array $\var{last}[1..p]$ where 
each process writes the index of the last block in
\var{root} containing an enqueue whose element it dequeued.
To perform GC, a process reads $\var{last}[1..p]$ and finds the maximum entry $m$.
Then, it helps complete every other process's pending dequeue 
by computing the dequeue's response and writing it in the block in the leaf that represents the dequeue.
Once this helping is complete, it follows from the FIFO property of the queue that elements enqueued 
in $\var{root}.\fld{blocks}[1..m-1]$ have all been dequeued, so GC can discard all blocks in the root's RBT 
with indices smaller than $m$.
Fortunately, there is an efficient \op{Split} operation that can remove
all nodes with indices less than $m$ from an RBT in logarithmic time \cite[Sec.~4.2]{Tar83}.
Then, GC can work its way down the \ordering\ tree, using \op{Split} operations to discard all subblocks
of the discarded root blocks.

>>>>

Known lock-free search trees have step complexity that includes a term linear in $p$ \cite{EFHR14,Ko20}.  
However, we do not require all the standard search tree operations.
Instead of a standard insertion, we allow a \op{Refresh}'s insertion to fail if another
concurrent \op{Refresh} succeeds in inserting a node, just as the CAS on line \ref{cas}
can fail if a concurrent \op{Refresh} does a successful CAS.
Moreover, the insertion should succeed only if the newly inserted block has a larger index than any other node in the RBT.
Thus, we can use a particularly simple concurrent RBT implementation.
A sequential RBT can be made persistent using the classic node-copying technique of 
Driscoll et al.~\cite{DSST89}:  all nodes are immutable, and operations on the 
tree are performed by making a new copy of each node $v$ that must be modified, as well
as each node along the path from the root to $v$.
The tree reachable from the new copy of the root is the result of applying the tree operation.
This only adds a constant factor to the running time of any routine designed for a (sequential) RBT.
Once a process has performed an update to the RBT representing the blocks of a node 
$v$ in the \ordering\ tree, 
it uses a CAS to swing $v$'s pointer from the previous RBT root to the new RBT root.
A search in the RBT can simply read the pointer to the RBT root and perform a standard
sequential search on it.
Bashari and Woelfel ~\cite{DBLP:conf/podc/BashariW21} used persistent red-black trees in a similar way to implement a snapshot data structure.

@@

To maintain good amortized time per operation, we periodically do a garbage collection (GC) phase.
If a process performing a \op{Refresh} on the root finds that $\var{root}.\head$ is a multiple
of $p^2$, it does GC.
To determine which blocks can be thrown away, we keep an array $\var{last}[1..p]$ where 
each process writes the index of the last block in
\var{root} containing an enqueue whose element it dequeued.
To perform GC, a process reads $\var{last}[1..p]$ and finds the maximum entry $m$.
Then, it helps complete every other process's pending dequeue 
by computing the dequeue's response and writing it in the block in the leaf that represents the dequeue.
Once this helping is complete, it follows from the FIFO property of the queue that elements enqueued 
in $\var{root}.\fld{blocks}[1..m-1]$ have all been dequeued, so GC can discard all blocks in the root's RBT 
with indices smaller than $m-1$.
Fortunately, there is an efficient \op{Split} operation that can remove
all nodes with indices smaller than $m-1$ from a RBT in logarithmic time \cite[Sec.~4.2]{Tar83}.
Then, GC can work its way down the \ordering\ tree, using \op{Split} operations to discard all subblocks
of the discarded root blocks.

### 2023-01-07 podc_paper/correctness.tex
working on GC

\begin{lemma} \label{lem::subblocksDistinct}
If $B$ and $B'$ are two blocks in nodes at the same depth, their sets of subblocks are disjoint.
\end{lemma}
\begin{proof}
We prove the lemma by reverse induction.
If $B$ and $B'$ are in leaves, they have no subblocks, so the claim is true.
Assume the claim is true for nodes at depth $d+1$ and let $B$ and $B'$ be two blocks in nodes at depth $d$.
Consider the direct subblocks of $B$ and $B'$ defined by (\ref{defsubblock}).
If $B$ and $B'$ are in different nodes at depth $d$, then their direct subblocks are disjoint.
If $B$ and $B'$ are in the same node, it follows from \Cref{lem::headProgress} that their direct subblocks are disjoint.
Either way, their direct subblocks (at depth $d+1$) are disjoint, so the claim follows from the induction hypothesis.
\end{proof}

>>>>

\begin{lemma} \label{lem::subblocksDistinct}
If $B$ and $B'$ are two blocks in nodes at the same depth, their sets of subblocks are disjoint.
\end{lemma}
\begin{proof}
We prove the lemma by reverse induction on the depth.
If $B$ and $B'$ are in leaves, they have no subblocks, so the claim holds.
Assume the claim holds for nodes at depth $d+1$ and let $B$ and $B'$ be two blocks in nodes at depth $d$.
Consider the direct subblocks of $B$ and $B'$ defined by~(\ref{defsubblock}).
If $B$ and $B'$ are in different nodes at depth $d$, then their direct subblocks are disjoint.
If $B$ and $B'$ are in the same node, it follows from \Cref{lem::headProgress} that their direct subblocks are disjoint.
Either way, their direct subblocks (at depth $d+1$) are disjoint, so the claim follows from the induction hypothesis.
\end{proof}
